---
title: "Checking language equivalence in R"
---

**NOTE: ALL FUNCTIONS HERE ARE SUPERSEDED BY THE k3 PACKAGE!**

R has a profound way of dealing with the unknown. `NA` represents missing values within our data, so that we can take them into account when operating on those data.

There is an entire logical system behind this, called [Kleene logic](https://en.wikipedia.org/wiki/Three-valued_logic#Kleene_and_Priest_logics). Any operation performed on `NA` will return `NA` if the result depends on the unknown value that `NA` stands for. This can lead to a highly counterintuitive result when checking for missing data:

```{r}
NA == NA
```

What happened? Because `==` follows Kleene logic, it asks whether the unknown values behind the two `NA`s are equal. We don't know these values, so we don't know the answer! If you're still confused, *R for Data Science* has an [nice example](https://r4ds.hadley.nz/logicals#sec-na-comparison).

This behavior makes perfect sense when you think about it. But there is still something unsatisfying here. Just looking at the syntax, the objects on both sides of `==` are the same. There must be a way to check for this!

And there is — we can use `identical()`. This function doesn't care about any hidden secrets. It just looks at the language objects and realizes that they are both `NA`, so it treats them as equal:

```{r}
identical(NA, NA)
```

In other words, `identical()` checks for *language equivalence*, not equality of the data that are conceptually represented by the language. Both objects are `NA`, so they are equivalent: we can swap one for the other, and the code will evaluate just the same.

This works in simple cases, but it doesn't scale: `identical()` is not vectorized, so it just checks whether each argument as a whole is the same as the other one. If we want to check each pair in a vector for language equivalence, we will be disappointed because we still only get a length-1 summary result:

```{r}
vec1 <- c(2, 4, NA, 7)
vec2 <- c(2, 5, NA, NA)
identical(vec1, vec2)
```

The most obvious solution would be to map `identical()` in a higher-order function or a for loop:

```{r}
library(purrr)
map2_lgl(vec1, vec2, identical)

# Using base R instead
mapply(identical, vec1, vec2, USE.NAMES = FALSE)

vecs_are_equal <- logical(length(vec1))
for (i in seq_along(vecs_are_equal)) {
  vecs_are_equal[i] <- identical(vec1[i], vec2[i])
}
vecs_are_equal
```

Yet `identical()` is a little strict. It's one of the very few functions that take the types of missing values into account:

```{r}
vec2
vec3 <- c("a", NA, NA, "z")

# Both are  missing...
vec2[3]
vec3[3]

# ...but not of the same type:
identical(vec2[3], vec3[3])

map2_lgl(vec2, vec3, identical)
```

This can also be a problem when comparing known values, especially when one vector is double and the other is integer. You likely don't care about the types of these vectors, only about their numeric values.

```{r}
vec1

# First vector is double, second is integer
map2_lgl(vec1, as.integer(vec1), identical)
```

Compare `identical()` to `==` which silently coerces the integer vector to double, so it doesn't see any differences between the known values:

```{r}
vec1 == as.integer(vec1)
```

What if we want this behavior for both known and missing values, so that all elements of the output would be `TRUE` here? Enter `is_equivalent()`:

```{r}
is_equivalent <- function(x, y, strict = FALSE) {
  if (strict) {
    return(mapply(identical, x, y, USE.NAMES = FALSE))
  }
  (is.na(x) & is.na(y)) | (!is.na(x) & !is.na(y) & x == y)
}
```

It checks each pair for language equivalence by default. For example, `2` and `as.integer(2)` count as equivalent, and so do any two missing values.

```{r}
is_equivalent(vec1, as.integer(vec1))
```

The difference between double and integer doesn't matter here because R mostly coerces the difference away. Indeed, for comparing two known values, `is_equivalent()` uses `==`. In general, the default behavior is:

-   Two known values are checked for equality using `==`.

-   Two missing values are equivalent, so the comparison returns `TRUE`.

-   One known and one missing value are not equivalent, so the comparison returns `FALSE`.

A faster version that applies this logic to two length-1 values would be:

```{r}
is_equivalent_scalar <- function(x, y, strict = FALSE) {
  if (strict) {
    return(identical(x, y))
  }
  (is.na(x) && is.na(y)) || (!is.na(x) && !is.na(y) && x == y)
}

# Better readable but less elegant:
is_equivalent_scalar2 <- function(x, y, strict = FALSE) {
  if (strict) {
    identical(x, y)
  } else if (is.na(x) && is.na(y)) {
    TRUE
  } else if (is.na(x) || is.na(y)) {
    FALSE
  } else {
    x == y
  }
}
```

As an alternative to this logic, we can set `strict` to `TRUE`, which will apply `identical()` pairwise. In other words, it makes the function behave like a vectorized `identical()`, checking each pair for exact equality:

```{r}
is_equivalent(vec1, as.integer(vec1), strict = TRUE)
```
