---
title: "Replace names in a file using R"
author: "Lukas Jung"
date: "2024-03-18"
categories: [code, analysis]
draft: true
---

You likely know about search and replace in RStudio. But what if you have many terms to replace? It would be very cumbersome and error-prone to do all of that manually. Here is how to automate it using R. The post offers you two dependency-free functions.

`replace_in_file()` takes a file path, a pattern to look for in that file, and a replacement for the pattern. It's the normal replace-all use case. But `replace_from_df()` scales this to any number of variables to replace.

You first need to collect all the names in a data frame: one column with the old names and one with the new ones. Besides a path argument, `replace_from_df()` also takes that data frame (`df_names`) as well as the names of two columns that store the old and new names (`col_pattern` and `col_replacement`) in the data frame. Further options are passed down to `gsub()`, which takes care of the actual replacement.

Each row in the data frame contains an old name and a new name. `replace_from_df()` replaces all instances of the old name by the new one. Using a data frame makes sure the two vectors of names are matched in length.

```{r}
replace_in_file <- function(path, pattern, replacement, ignore_case = FALSE,
                            fixed = FALSE, use_bytes = FALSE) {
  # Read the content of the file into a variable:
  file_content_old <- readLines(path)

  # Replace the word in the content:
  file_content_new <- gsub(
    pattern = pattern,
    replacement = replacement,
    x = file_content_old,
    ignore.case = ignore_case,
    fixed = fixed,
    useBytes = use_bytes
  )

  # Write the updated content back to the file:
  writeLines(file_content_new, path)
}

# Function that goes to the file at `path` and takes `df_names` as a lookup
# table to replace the values in the `col_pattern` column by those in
# `col_replacement`:
replace_from_df <- function(path, df_names, col_pattern, col_replacement,
                            ignore_case = FALSE, fixed = FALSE,
                            use_bytes = FALSE) {
  if (!all(c(col_pattern, col_replacement) %in% colnames(df_names))) {
    stop("`col_pattern` and `col_replacement` must be column names of `df_names`.")
  }
  for (i in seq_len(nrow(df_names))) {
    replace_in_file(
      path = path,
      pattern = df_names[[col_pattern]][i],
      replacement = df_names[[col_replacement]][i],
      ignore_case = ignore_case
    )
  }
}
```

I recommend `tibble::tribble()` for creating the data frame. In this way, each pair of names is in the same row:

```{r}
cakes <- tibble::tribble(
  ~messy,                      ~tidy,
  "My Creamy Apple Pie!",      "apple_pie",
  "Yummy Angel Food Cake...",  "angel_food_cake",
  "Black Forest cake?!",       "black_forest_cake"
)
```

Now replace all names in your file. It's not a very realistic example, but you get the idea.

```{r}
replace_from_df(
  path = "recipes/cakes.R",
  df_names = cakes,
  col_pattern = "messy",
  col_replacement = "tidy"
)
```
