---
title: "How to find the mode in R (or not)"
format:
  html:
    code-fold: false
    code-summary: "Show code"
---

R is made for statistics. Surprisingly, though, there is no built-in R function to find the statistical mode: the most frequent value in a given distribution. (The base R function `mode()` has a very different use case.)

Here are three mode functions for you.

All functions behave the same way if there is only one mode. If there are multiple modes, `mode_first()` only returns the one that appears first. `mode_all()` returns all modes. `mode_single()` returns `NA` in this case.

```{r}
mode_first <- function(x, na.rm = FALSE) {
  if (na.rm) {
    x <- x[!is.na(x)]
  }
  ux <- unique(x[!is.na(x)])
  mode1 <- ux[which.max(tabulate(match(x, ux)))]
  if (any(is.na(x))) {
    adjudicate_mode_na(x, ux, mode1, TRUE)
  } else {
    mode1
  }
}

mode_all <- function(x, na.rm = FALSE) {
  if (na.rm) {
    x <- x[!is.na(x)]
  }
  ux <- unique(x[!is.na(x)])
  tab <- tabulate(match(x, ux))
  modes <- ux[tab == max(tab)]
  if (length(modes) == 1L) {
    adjudicate_mode_na(x, ux, modes, FALSE)
  } else if (any(is.na(x))) {
    methods::as(NA, typeof(x))
  } else {
    modes
  }
}

mode_single <- function(x, na.rm = FALSE) {
  if (na.rm) {
    x <- x[!is.na(x)]
  }
  mode1 <- mode_all(x, na.rm)
  if (length(mode1) == 1L) {
    adjudicate_mode_na(x, unique(x[!is.na(x)]), mode1, FALSE)
  } else {
    methods::as(NA, typeof(x))
  }
}

# Helper used in all functions above; don't call directly:
adjudicate_mode_na <- function(x, ux, mode1, allow_tie) {
  mode2 <- ux[which.max(tabulate(match(x[x != mode1], ux)))]
  count_mode1 <- length(x[x == mode1])
  count_mode2_na <- length(x[x == mode2]) + length(x[is.na(x)])
  if (allow_tie) {
    mode1_frequent_enough <- count_mode1 >= count_mode2_na
  } else {
    mode1_frequent_enough <- count_mode1 > count_mode2_na
  }
  if (mode1_frequent_enough) {
    mode1
  } else {
    methods::as(NA, typeof(x))
  }
}
```

At their core, these functions are based on a [Stack Overflow answer by Ken Williams](https://stackoverflow.com/questions/2547402/how-to-find-the-statistical-mode/8189441#8189441). Go to *Implementation notes* below for details.

What is so special about these functions? They really care about `NA`s. Let's have a look.

## Get the first mode with `mode_first()`

Everything is fine here:

```{r}
x1 <- c(7, 8, 8, 9, 9, 9)
mode_first(x1)
```

But what if some values are missing? Think about what it means to look for the most frequent value in such a distribution. Maybe there are so many missings that it's impossible to tell which value is the most frequent one. For example:

```{r}
x2 <- c(1, 1, 2, 2, 2, 2, NA, NA, NA, NA)
mode_first(x2)
```

If each `NA` stands in for `1`, then `1` is the most frequent value. Accordingly for `2`. The mode of `x1` depends on the true values hiding behind `NA`. Since we don't know these values, we don't know the mode! The function should return `NA`, and it does.

Ignore `NA`s using `na.rm = TRUE` if you must:

```{r}
mode_first(x2, na.rm = TRUE)
```

This distribution is different:

```{r}
x3 <- c(7, 7, 8, 8, 8, 8, NA)
mode_first(x3)
```

Even if the `NA` stands in for `7`, there will only be three instances of `7` but four instances of `8`. The mode is `8`, independent of the true value behind `NA`.

## Get all modes with `mode_all()`

This function captures multiple modes:

```{r}
x4 <- c("a", "a", "b", "b", "c", "d", "e")
mode_all(x4)
```

If some values are missing but there would be multiple modes when ignoring `NA`s, `mode_all()` returns `NA`. That's because missings can easily create an imbalance between the equally-frequent known values:

```{r}
x5 <- c(1, 1, 2, 2, NA)
mode_all(x5)
```

If `NA` masks either `1` or `2`, that number is the (single) mode. As before, if the mode depends on missing values, the function returns `NA`.

Yet `na.rm = TRUE` makes the function ignore this:

```{r}
mode_all(x5, na.rm = TRUE)
```

## Get the single mode (or `NA`) with `mode_single()`

`mode_single()` is stricter than `mode_first()`: It returns `NA` if there are --- or might be --- multiple modes. Otherwise, it works the same way.

```{r}
x6 <- c(3, 4, 4, 5, 5, 5)
mode_single(x6)

x7 <- c("x", "x", "y", "y", "z")
mode_single(x7)

x8 <- c(1, 1, 2, NA)
mode_single(x8)
```

```{r eval=FALSE, include=FALSE}

# COMPARE IMPLEMENTATIONS:

x <- list(x1, x2, x3, x4, x5, x6, x7, x8)

# The two original functions by Ken Williams:
Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}
Modes <- function(x) {
  ux <- unique(x)
  tab <- tabulate(match(x, ux))
  ux[tab == max(tab)]
}

df <- tibble::tibble(
    x = x,
    mode_first = unlist(purrr::map(x, mode_first)),
    mode_single = unlist(purrr::map(x, mode_single)),
    Mode = unlist(purrr::map(x, Mode)),
    mode_all = purrr::map(x, mode_all),
    Modes = purrr::map(x, Modes)
)

df
```

## Implementation notes

The first two functions are adapted with modifications from [this Stack Overflow answer](https://stackoverflow.com/questions/2547402/how-to-find-the-statistical-mode/8189441#8189441). They only behave differently from the original ones when handling missing values. I changed their names to better distinguish them from `base::mode()` and from each other. Note that all three functions internally call the helper `adjudicate_mode_na()`.

The `na.rm` argument follows similar R functions, such as `mean()` and `median()`. Still, it shouldn't be set to `TRUE` without a strong rationale. Missing values have real meaning, and all these functions handle them accordingly --- like [base R operators do](https://adv-r.hadley.nz/vectors-chap.html#missing-values).

Click "Show code" to see step-by-step commented versions of all three functions.

```{r eval=FALSE}
#| code-fold: true

mode_first <- function(x, na.rm = FALSE) {
  # We first determine the unique known values
  # of `x`. `NA`s are ignored at this point
  # because they will receive special treatment
  # later on.
  ux <- unique(x[!is.na(x)])
  # Find the index of the most frequent known
  # value (the possible mode) and subset this
  # value:
  mode1 <- ux[which.max(tabulate(match(x, ux)))]
  # The present implementation only differs from the
  # linked original function in terms of `NA` handling.
  # Therefore, it returns `mode1` just like that function
  # does if missing values are not an issue -- either
  # because the user chose to ignore them (`na.rm = TRUE`)
  # or because there are no `NA`s to begin with:
  if (na.rm || !any(is.na(x))) {
    mode1
  } else {
  # At this point, we know that one or more
  # values are missing. We call a helper function,
  # documented further below, that adjudicates
  # whether or not it's still possible to determine
  # the mode. If so, it returns the mode; if not,
  # it returns `NA`:
    adjudicate_mode_na(x, ux, mode1, TRUE)
  }
}

mode_all <- function(x, na.rm = FALSE) {
  # As above, we first determine the unique
  # known values of `x`. `NA`s are ignored
  # at this point because they will receive
  # special treatment later on.
  ux <- unique(x[!is.na(x)])
  # Count the instances of each known value:
  tab <- tabulate(match(x, ux))
  # Subset the vector of unique known values
  # at the indices corresponding to the
  # most frequent known values:
  modes <- ux[tab == max(tab)]
  # A seemingly unimodal distribution is
  # subject to the `NA`-related caveats
  # described in `mode_first()`, so we call
  # the same `NA` helper as that function:
  if (length(modes) == 1L) {
    adjudicate_mode_na(x, ux, modes, TRUE)
    # Multimodal distributions without `NA`s
    # have a clearly determined set of modes.
    # If they do have `NA`s, the user can
    # (but probably shouldn't) ignore the
    # issue by setting `na.rm` to `TRUE`:
  } else if (na.rm || !any(is.na(x))) {
    modes
    # Any missing value could mask any of the
    # known values tied for most frequent --
    # and break the tie. This makes it
    # impossible to determine the true set
    # of modes, so the function returns `NA`:
  } else {
    methods::as(NA, typeof(x))
  }
}

mode_single <- function(x, na.rm = FALSE) {
  # Unlike the functions above, this
  # one requires manual `NA` removal:
  if (na.rm) {
    x <- x[!is.na(x)]
  }
  # We also need to check the number of
  # modes here, so we call `mode_all()`:
  mode1 <- mode_all(x, na.rm)
  # As the name says, if the distribution
  # has a single mode (that passes the
  # `NA` test), that value is returned.
  # `NA` testing without allowing for ties
  # between the `mode1` count and the sum
  # of the `mode2` and `NA` counts
  # (`FALSE` at the end) is necessary here
  # because we need to make sure that
  # `mode1` is really the only mode, even
  # if all `NA`s stand in for `mode2`:
  if (length(mode1) == 1L) {
    adjudicate_mode_na(x, unique(x[!is.na(x)]), mode1, FALSE)
  # Multimodal distributions are always `NA`.
  # Some users prefer this stricter way of
  # estimating the mode, or they require it
  # for their specific use cases.
  } else {
    methods::as(NA, typeof(x))
  }
}

adjudicate_mode_na <- function(x, ux, mode1, allow_tie) {
  # Some values might be missing. We need to check
  # whether there are so many missings that the most
  # frequent known value, `mode1`, might be less
  # frequent than the second-most frequent one (or
  # the first value tied with `mode1` but appearing
  # after it) if all the `NA`s mask the latter.
  # To do so, we need to find this second value.
  # We look for a possible mode much like above,
  # but this time, we exclude the `mode1` values:
  mode2 <- ux[which.max(tabulate(match(x[x != mode1], ux)))]
  # Count instances of the three relevant available
  # values -- most and second-most frequent known
  # values plus missing values:
  count_mode1 <- length(x[x == mode1])
  count_mode2_na <- length(x[x == mode2]) + length(x[is.na(x)])
  # Two of the functions above only require `mode1`
  # to be at least as frequent as the sum of the
  # other two counts (see below). However,
  # `mode_single()` is more strict because
  # it is meant to rule out that the true count
  # of `mode2` is just as high as that of `mode1`.
  # If this is even possible, `mode_single()`
  # needs to return `NA` because it only
  # allows for a single mode.
  if (allow_tie) {
    mode1_frequent_enough <- count_mode1 >= count_mode2_na
  } else {
    mode1_frequent_enough <- count_mode1 > count_mode2_na
  }
  # (Assuming `allow_tie = TRUE`:)
  # `mode1` is the true mode only if it's
  # at least as frequent as the sum of the
  # other two counts. Otherwise, if all the
  # `NA`s mask `mode2` values, the true count
  # of `mode2` would be higher than that of
  # `mode1`. We don't know which values hide
  # behind `NA`, so we can't rule out this
  # second scenario if the known count of
  # `mode1` is lower than a possibly true
  # count of `mode2`. Therefore, if
  # `count_mode1` is not large enough, the
  # function returns `NA` (coerced to the
  # same type as the input, `x`):
  if (mode1_frequent_enough) {
    mode1
  } else {
    methods::as(NA, typeof(x))
  }
}
```

## Learn more

What about other mode estimation techniques in R? Search this [CRAN Task View](https://cran.r-project.org/web/views/Distributions.html) for "Mode estimation". However, I haven't checked how the packages listed there deal with missing values.
